Distribution-oblivious Programming with Dish

### 2019-11-21 pldi/draft.tex
A draft of the front end

@@

In order to address this issue, we introduce the notion of
distributable program regions, and we design an analysis that given the
abstract syntax tree of a shell script identifies distributable
regions and translates each of them to a dataflow graph. The
AST of the shell script is produced using a POSIX compliant parser
\kk{Reference libdash github. This reference should probably go to
  evaluation}.

@@

\kk{I wanted this to be purely about the regions, and a next
  subsection be about the analysis, but the separation line is very
  blurry, so they are now in the same subsection.}

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 &&
  sort f3 #Replace
\end{lstlisting}

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \texttt{cat f1 f2} and
\texttt{grep ``foo''} already execute concurrently in the standard
shell, while \texttt{sort f3} waits for their completion before being
executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and \texttt{sort
  f3} are maximal distributable regions, in the sense that they
cannot be extended beyond \texttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following shell constructs: ;, \&\&, \texttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \texttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

@@

Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom
up, translating their \todo{independent} components to dataflow graph
nodes, until it reaches a barrier construct. Note that barrier
constructs can be contained in a distributable subexpression, as a
\todo{whole independent component}. An illustrative example follows:

@@

In the above example\kk{Can we make example not floating?}, while
\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
  g2} are all in different distributable regions. The dataflow graph
that corresponds to the \todo{large} region, has two nodes,
\texttt{sort f1 > g1 \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.

The analysis algorithm also finds the category of each node that it
adds to the dataflow graph, using the rules described in \ref{}. Some
interesting cases of the analysis algorithm are shown in pseudocode
below:

\begin{lstlisting}[language=python, float=h]
  def analysis(node):
    ...
    if(node.name == 'Pipe'):
      return pipe_graphs([analysis(child)
        for child in node.children])
    elif(node.name == 'And'):
      node.children = [analysis(child)
        for child in node.children]
      return node
    elif(node.name == 'Command'):
      if(not safe(node)):
        return make_ir_node(node, 'unsafe')
      else:
        return make_ir_node(node,
          find_category(node))
    ...
\end{lstlisting}

\kk{Explain how the three above cases work, mentioning what does
  pipe-graphs and safe do. The important point to get across is to
  explain why we do the ``conservative checks'' in safe, and how they
  preserve safety.}

\paragraph{analysis}

On the other hand: A distributable subtree starts whenever there is a
pipe, or asynchronous list. If the subcomponents of the pipe or
asynchronous list were already graphs, then we connect them to the
bigger graph \kk{explain how we do pipe and asynchronous list}, and if
they were not graphs, we make each subcomponent a node in the subtree,
event though they are not commands. These nodes do not belong to any
category, but still have stdin and stdout, and can be executed
concurrently (as long as they satisfy the checked assumptions above),
since they don't introduce new behaviours, they would still be
executed concurrently when ran in the standard shell, producing
possibly non-deterministic results with different interleavings.

\kk{Explain how we merge two graphs when seeing a pipe}

Whenever we see a subtree of the AST that is a command, we check the
files that it writes to. If any of its arguments is a command
substitution or contains anything that is evaluated\kk{make this more
  precise}, its category is marked as sequential\kk{name?}. Otherwise
we check its its category and its input output arguments (using the
methods described in \Cref{ext}). The command is then initialized as
the node of a singular graph, with the inputs and outputs of the graph
being the ones of the node.

@@

\kk{Rephrase that paragraph, to mention the assumtpions that the front
  end makes, and why they are not so unreasonable}

There are some assumptions that we make in order to preserve the
correctness of the computation under distributability optimizations.

\kk{Our notion of distributable subtrees has two dimensions. First,
  the state of the shell should be an invariant in a distributable
  subtree. We cannot know this completely without evaluating, so we do
  a conservative analysis. The second dimension is that there exists
  at least one component that can be abstracted as a node of the
  dataflow graph. This can be anything that includes a command.}

%% Even though commands are split in categories, and several pipelines
%% are distributable, not all parts of a shell script are stateless and
%% several of them have effects that could affect the behaviour of a
%% program after distribution.

\kk{All of our assumptions and study are based on the POSIX
  standard. Reference?}

\kk{Not checked Assumption. A command doesn't write/read to arbitrary
  files that are not mentioned in its arguments. This is necessary to
  enforce that no two commands concurrently read/write from the same
  file.}

\kk{Not checked Assumption: No signals, traps, and sets at the
  moment. At the moment we don't have a way to handle signals and
  traps.}

\kk{Partially checked Assumption. Only one reader and one writer to
  each file, and no cycles. OLD MORE CONSERVATIVE ASSUMPTION: The same
  file exists once in the substree. This is a conservative assumption
  that ensures that no two commands write or read to the same file
  concurrently. We kind of check it but not completely. We check that
  the expanded evaluated files are not the same, but one could have
  part of a file in a variable etc.}

\kk{Checked assumption: How can we make sure that nodes in the IR are
  deterministic. Do we make some assumption that enforces this? I
  guess that commands that are nodes in the IR are all deterministic?
  That is stateless and pure commands are deterministic.}

>>>>

\kk{The writing in this section is wandering around without a clear
  purpose :( }

@@

In order to address this issue, we introduce the notion of
distributable program regions, and we design a translation pass that
given the abstract syntax tree of a shell script identifies
distributable regions and translates each of them to a dataflow
graph. The AST of the shell script is produced using a POSIX compliant
parser \kk{Reference libdash github. This las sentence should probably
  go to evaluation}.

@@

%% \kk{I wanted this to be purely about the regions, and a next
%%   subsection be about the analysis, but the separation line is very
%%   blurry, so they are now in the same subsection.}

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 &&
  sort f3 # Replace with sth better
\end{lstlisting}

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \texttt{cat f1 f2} and
\texttt{grep ``foo''} would already execute concurrently in the
standard shell, while \texttt{sort f3} waits for their completion
before being executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and
\texttt{sort f3} are maximal distributable regions, in the sense that
they cannot be extended beyond \texttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following barrier constructs: ;, \&\&, \texttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \texttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

@@

\subsection{Translation Pass}

Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom up,
translating their independent components to dataflow graph
nodes, until a barrier construct is reached. Note that barrier
constructs can be contained in a dataflow graph, as long as they don't
extend beyong a single node. An illustrative example follows:

@@

In the above example\kk{Can we make example not floating?}, while
\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
  g2} are all in different dataflow graphs. The dataflow graph that
corresponds to the whole program, has two nodes, \texttt{sort f1 > g1
  \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.

In order to identify possible opportunities for parallelism, the
translation pass also identifies the category of each node, together
with its input and output, using the rules described in \ref{}. Some
interesting cases of the analysis algorithm are shown in pseudocode
below:

\begin{lstlisting}[language=python, float=h]
  def translate(node):
    ...
    elif(node.name == 'Pipe'):
      return pipe_graphs([translate(child)
        for child in node.children])
    elif(node.name == 'And'):
      node.children = [translate(child)
        for child in node.children]
      return node
    elif(node.name == 'Command'):
      if(not safe(node)):
        return make_ir_node(node, 'unsafe')
      else:
        return make_ir_node(node,
          find_category(node))
    ...
\end{lstlisting}

%% \kk{Explain how the three above cases work, mentioning what does
%%   pipe-graphs and safe do. The important point to get across is to
%%   explain why we do the ``conservative checks'' in safe, and how they
%%   preserve safety.}

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph \kk{Does this need an
  example?}. Whenever the algorithm encounter an \texttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then doesn't merge them in a large dataflow graph. Finally,
whenever the algorithm encounters a command, it first checks whether
it is safe to possibly parallelize this command. A command is
considered to be safe, if it does not make any modification to the
shell state (e.g. changing the current directory, modifying
environment variables), and if it does not read from a
\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
  unsafe command can still be safely added to the dataflow graph as a
  single node, since this doesn't introduce any new
  behaviours. However it cannot be parallelized further than that.} If
the command is considered safe, its category is identified and a node
containing that command is created and added to a singleton dataflow
graph.

It is important to note that due to the highly dynamic nature of the
shell, the safety properties that were mentioned above cannot be
soundly checked statically. For example in the first case, two
unexpaned strings could refer to the same file, but the translation
pass cannot infer that. Hence, in both cases, the checks were
approximated using a \todo{best-effort} method. \kk{Maybe mention that
  as future work, we want to tighly integrate with the shell shtepper
  that will allow to maximally expand everything before making this
  decision.}

@@

\kk{This paragraph is not extremely necessary. Maybe it can be moved
  to limitations?}

There are two additional assumptions that have to be satisfied in
order for the distributed implementation that is generated by Dish to
have equivalent behaviour with the original implementation. First of
all, there should be no external signals to the commands while they
execute. At the moment Dish assumes no faults, and that communication
always succeeds. This assumption can be lifted by extending Dish with
a fault-tolerant runtime, \kk{blah blah}. Second, we assume that no
command reads and writes to arbitrary files that are not mentioned in
their arguments. This is a necessary assumption, since commands are
considered as black boxes and the only information that can be
inferred about them is through their arguments.

\kk{The following assumption should probably be mentioned in the
  categories, since a command that is non-deterministic should not be
  categorized to be neither stateless or pure. Assumption: How can we
  make sure that nodes in the IR are deterministic. Do we make some
  assumption that enforces this? I guess that commands that are nodes
  in the IR are all deterministic?  That is stateless and pure
  commands are deterministic.}

%% \kk{All of our assumptions and study are based on the POSIX
%%   standard. Reference?}
